---
title: "Head-to-head: An interactive d3 map"
author: "Piotr Patrzyk"
output: 
  html_document:
    keep_md: false
---

```{r setup, include=FALSE}
library(data.table)
library(r2d3)
library(jsonlite)
library(DT)

knitr::opts_chunk$set(
  echo = FALSE,
  out.width = '100%'
)
```

```{r customcss, results="asis"}
cat("
<style>
svg {
  display: block;
  margin: auto;
}

.tooltip {
  position: absolute;
  text-align: center;
  padding-left: 10px;
  padding-right: 10px;
  padding-top: 3px;
  padding-bottom: 3px;
  background: #d4d6d8;
  border: 0px;
  border-radius: 8px;
  pointer-events: none;
}

.country-border {
  stroke: white;
  stroke-width: 0.5;
}
</style>
")
```

# Introduction

In this kernel I visualize head-to-head balances in international footbal games. For any given team, you can look up:

- what other teams have they played against,
- with whom do they usually win or lose.

## Dataset

```{r dataload}
games <- fread('results.csv', encoding = 'UTF-8')
# fix necessary to match geojson
games[home_team == 'Korea Republic', home_team := 'South Korea']
games[away_team == 'Korea Republic', away_team := 'South Korea']
games[home_team == 'Korea DPR', home_team := 'North Korea']
games[away_team == 'Korea DPR', away_team := 'North Korea']
```

The dataset contains `r games[, .N]` matches played by `r games[, uniqueN(c(home_team, away_team))]` different teams. It is aggregated to obtain head-to-head summary (#wins, #draws, #losses) for all team pairs in the dataset.

```{r getresults}
games[, goal_diff := home_score - away_score]
unique_teams <- games[, unique(c(home_team, away_team))]
# outcomes_list is in r2d3-friendly format, headtoheadfor analysis in R
outcomes_list <- headtohead <- replicate(length(unique_teams), data.table())
names(outcomes_list) <- unique_teams
for (i in 1:length(unique_teams)) {
  current_team <- unique_teams[i]
  current_games <- games[home_team == current_team | away_team == current_team, ]
  current_games[, location_home := (home_team == current_team)]
  current_games[, rival := ifelse(location_home, away_team, home_team)]
  current_games[location_home == FALSE, goal_diff := -goal_diff]
  outcomes <- current_games[,
    .(
      team = current_team,
      win = sum(goal_diff > 0),
      draw = sum(goal_diff == 0),
      loss = sum(goal_diff < 0),
      total = .N
    ),
    by = rival
  ]
  outcomes[, difference := (win - loss)]
  outcomes[, metric := round(difference / total, 2)]
  tooltip <- outcomes[, sprintf('%s vs. %s<br>W:%s D:%s L:%s', current_team, rival, win, draw ,loss)]
  metric <- outcomes[, metric]
  total <- outcomes[, total]
  names(metric) <- names(tooltip) <- names(total) <- outcomes[, rival]
  headtohead[[i]] <- outcomes[, .(team, rival, win, draw, loss, difference, total, metric)]
  outcomes_list[[i]] <- list(
    metric = as.list(metric),
    total = as.list(total),
    tooltip = as.list(tooltip)
  )
}
headtohead <- rbindlist(headtohead)
# toJSON(outcomes_list, auto_unbox = TRUE)
```

There are `r headtohead[, .N/2]` unique pairs in the dataset. Given that there are `r choose(games[, uniqueN(c(home_team, away_team))], 2)` possible pairs, this means that only `r paste0(round(100 * headtohead[, .N/2] / choose(games[, uniqueN(c(home_team, away_team))], 2), 2), "%")` of possible matches actually took place.

Top 10 most often meeting teams:

```{r totalmax}
headtohead[order(-total), ][seq(1, 20, 2), .(team, rival, win, draw, loss, total)]
```

The metric that is displayed on the map below is obtained by using the formula: 

$$
metric = \frac{won-lost}{total\_played}
$$
The metric ranges from -1 (all games lost against given rival) to 1 (all games won). In case of even balance it equals to 0.

For example:

```{r scotland}
headtohead[team == 'Scotland' & rival == 'Wales', ]
```

# Map

Balances for a chosen team are displayed. Countries against which the balance is positive are colored in green, negative in red. In order to **change country** for which head-to-head stats are displayed **click on it** on the map. Zooming (wheel / double-click) and panning (drag) also work.

```{d3 map, data=outcomes_list}
var margin = {top: 25, right: 25, bottom: 25, left: 25};
var map_default_scale = 175.295;
var map_tooltip = null;
var map_json = null;
var color_scale = d3.scaleSequential(d3.interpolatePiYG);
var missing_color = "#000000";

var map = svg
  .attr("width", width + margin.left + margin.right)
  .attr("height", height + margin.top + margin.bottom)
  .attr("class", "border")
  .append("g")
  .attr("transform", "translate(" + margin.left + "," + margin.top + ")")
  .attr("id", "mapgroup");

function render_map(initial_country){
  
  color_scale.domain([-1, 1]);
  
  var projection = d3.geoNaturalEarth1()
    .translate([width/2, height/2])
    .scale(map_default_scale);

  var path = d3.geoPath().projection(projection);
  
  map_tooltip = d3.select("body").append("div")
    .attr("class", "tooltip")
    .style("opacity", 0);  
    
  map.append("rect")
    .attr("x", 0).attr("y", 0)
    .attr("width", width).attr("height", height)
    .attr('fill', '#cfeff9')
    .attr("opacity", 0);
  
  map.selectAll("path")
    .data(map_json.features)
    .enter()
    .append("path")
    .attr("d", path)
    .attr("class", "country-border")
    .style("fill", function(d) {
      var country = d.properties[initial_country];
      if (country) {
        return color_scale(country.metric);
      } else {
        return missing_color;
      }
    })
    .on("mouseover", function(d) {
      var content = d.properties[initial_country];
      if (typeof content !== 'undefined'){
        map_tooltip.transition().duration(200).style("opacity", 0.9);
        map_tooltip.html(content.tooltip)
          .style("left", (d3.event.pageX) + "px")
          .style("top", (d3.event.pageY - 28) + "px");
      }
    })
    .on("mouseout", function(d) {
      map_tooltip.transition().duration(500).style("opacity", 0);
    })
    .on("click", function(d) {
      update_map(d.properties.admin);
    });
    
  var zooming = function(d) {
    var offset = [d3.event.transform.x, d3.event.transform.y];
    var scale_update = (d3.event.transform.k * map_default_scale);
    projection.translate(offset)
      .scale(scale_update);
    map.selectAll("path")
      .attr("d", path);
  };
  
  var zoom_handler = d3.zoom()
    .scaleExtent([1, 20])
    .translateExtent([
      [-width/2+margin.left, -height/2+margin.top],
      [width/2+margin.right, height/2+margin.bottom]
    ])
    .on("zoom", zooming);
  
  map
    .call(zoom_handler)
    .call(zoom_handler.transform, function(){
      return d3.zoomIdentity.translate(width/2, height/2);
    })
    .on("wheel", function() {
      d3.event.preventDefault();
  });
    
  map.append("text")
    .attr("text-anchor", "start")
    .attr("font-weight", "bolder")
    .attr("x", 0)
    .attr("y", 0)
    .text('Display: '+initial_country);
};

function update_map(new_country){
  map.selectAll("path")
    .style("fill", function(d) {
      var country = d.properties[new_country];
      if (country) {
        return color_scale(country.metric);
      } else {
        return missing_color;
      }
    })
    .on("mouseover", function(d) {
      var content = d.properties[new_country];
      if (typeof content !== 'undefined'){
        map_tooltip.transition().duration(200).style("opacity", 0.9);
        map_tooltip.html(content.tooltip)
          .style("left", (d3.event.pageX) + "px")
          .style("top", (d3.event.pageY - 28) + "px");
      }
    })
    .on("mouseout", function(d) {
      map_tooltip.transition().duration(500).style("opacity", 0);
    })
    .on("click", function(d) {
      console.log(d.properties.admin);
      update_map(d.properties.admin);
    });
  map.select("text").text('Display: '+new_country);
};
  
d3.json('https://raw.githubusercontent.com/ppatrzyk/kaggle/master/football/map.json')
  .then(function(json){
    map_json = json;
    for (var country in data) {
      var metric = data[country]['metric'];
      var tooltip = data[country]['tooltip'];
      var total = data[country]['total'];
      // todo set scale
      for (var key in metric) {
        if (metric.hasOwnProperty(key)) {
          var current_metric = metric[key];
          var current_tooltip = tooltip[key];
          var current_total = total[key];
          for (var i = 0; i < map_json.features.length; i++) {
            var current_country = map_json.features[i].properties.admin;
            if (key == current_country) {
              map_json.features[i].properties[country] = {}
              map_json.features[i].properties[country].metric = current_metric;
              map_json.features[i].properties[country].tooltip = current_tooltip;
              map_json.features[i].properties[country].total = current_total;
              break;
            }
          }
        }
      }
    }
    render_map('Poland');
  });

```

<br>

# Not included on map

It's impossible (or at least impractical) to map overlapping polygons on choropleth map, so balances for some countries need to be presented separately. Games played by Soviet Union are counted for Russia in the dataset.

### Czechoslovakia

```{r czsk}
datatable(
  headtohead[
    team == 'Czechoslovakia',
  ][
    order(-total),
  ]
)
```

### Yugoslavia

```{r yugo}
datatable(
  headtohead[
    team == 'Yugoslavia',
  ][
    order(-total),
  ]
)
```

### East Germany

```{r nrd}
datatable(
  headtohead[
    team == 'German DR',
  ][
    order(-total),
  ]
)
```

# Limitations

Two warnings about using head-to-head summaries in team comparisons ("which team is better"):

Differences are not transitive. If team A has positive balance agsint team B and team B has positive balance against team C it does not necessarily mean that team A has positive balance against team C. For Instance, consider these head-to-head balances:

```{r frenedbel}
headtohead[
  (team == 'France' & rival == 'Netherlands') |
    (team == 'Netherlands' & rival == 'Belgium') |
    (team == 'France' & rival == 'Belgium'), 
]
```

Dataset contains historical games tracking back to 19th century and many teams performed differently over years. For instance, Hungary -- arguably not a top football team nowadays -- maintains a positive balance against many teams, including Brazil and France:

```{r hunexample}
headtohead[team == 'Hungary' & rival %in% c('Brazil', 'France'), ]
```

However, the last time they defeated France was in 1976 and all games since then were lost.

```{r hunbra}
games[
  as.Date(date) > as.Date('1976-01-01') &
  ((home_team == 'Hungary' & away_team == 'France') |
    (away_team == 'Hungary' & home_team == 'France')),
  .(date, home_team, away_team, home_score, away_score, city)
]
```