---
title: "Music Production Across the World"
author: "Piotr Patrzyk"
output: 
  html_document:
    keep_md: false
    toc: true
    toc_depth: 2
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(
  fig.align = 'center', 
  out.width = '80%',
  dev = 'png'
)
```

# Introduction

An exploratory analysis of data on music popularity. Where do the most popular artists come from? Which musical genres are the most popular? Are musical genres tied to specific countries? In this kernel, I look at the country of origin of musical artists and compare what types of music specific countries export.

## Preparing data

```{r helperfunctions, echo=FALSE}
display_values <- function(number){
  if(length(number) > 1){
    return(sapply(number, display_values))
  }
  if(is.na(number) | is.null(number)){
    number <- NA_character_
  }else if(number < 10^3){
    number <- as.character(round(number, 1))
  }else if(number < 10^6){
    number <- paste0(round(number / 10^3, 1), "k")
  }else if(number < 10^9){
    number <- paste0(round(number / 10^6, 1), "M")
  }else if(number < 10^12){
    number <- paste0(round(number / 10^9, 1),"B")
  }
  return(number)
}
```

```{r dataload, echo=FALSE, message = FALSE, warning = FALSE}
library(data.table)
library(foreach)
library(doParallel)
library(geojsonio)
library(ggplot2)
library(viridis)
library(hexbin)
library(DT)
library(leaflet)
library(wordcloud)

lfm <- fread('artists.csv', encoding = 'UTF-8', na.strings = '')
lfm[country_mb == "Soviet Union", country_mb := 'Russia']
lfm[country_mb == "East Germany", country_mb := 'Germany']
lfm[country_mb == "Czechoslovakia", country_mb := 'Czech Republic']
lfm[country_mb %in% c('Yugoslavia', 'Serbia and Montenegro'), country_mb := 'Serbia']
lfm[country_mb == 'Europe', country_mb := NA_character_]
```

First I read and prepare the data. Dataset has `r nrow(lfm)` rows and `r ncol(lfm)` columns. First row:

```{r examplerow, echo=FALSE}
t(lfm[1L, ])
```

Dataset needs to be cleaned; specifically, this entails:

- removing duplicates in artists,
- establishing single country of origin,
- establishing single genre from list of tags.

## Duplicated artist names

Due to the fact that last.fm does dot disambiguate between different artists that have the same name (see dataset description), dataset contains some artists with duplicated listeners/scrobbles count, coming from the aggregate profile. For example:

```{r duplicateexample, echo=FALSE}
lfm[
  artist_lastfm == 'First Aid Kit', 
  .(artist_lastfm, country_mb, country_lastfm, listeners_lastfm)
]
```

Let's investigate and quantify the problem. First these are the top 10 repeated artist names in the dataset (exact name match, case insensitive):

```{r duplicates, echo=FALSE}
name_duplicates <- lfm[,
  .(unique_artists = .N),
  by = .(name = tolower(artist_mb))
][
  order(unique_artists, decreasing = TRUE),
]
name_duplicates[, name := factor(name, levels = rev(name))]

duplicate_bar <- ggplot(
  name_duplicates[1:10, ],
  aes(name, unique_artists)
) +
  geom_col(
    fill = '#ED7953FF'
  ) +
  coord_flip() +
  labs(
    title = "Duplicate artist names (case insensitive)",
    x = "Name",
    y = "Unique artists"
  ) +
  theme_bw() +
  theme(
    axis.text.y = element_text(color = "black", size = 12)
  )

duplicate_bar
```

In general, `r paste0(round(100 * name_duplicates[unique_artists > 1, sum(unique_artists)] / name_duplicates[, sum(unique_artists)], 2), "%")` of all artists recorded in the dataset have a name that is not unique. That's a non-negligible proportion. On the one hand, this can be expected -- people have similar ideas and they will come up with exactly the same names independently. On the other, it's kind of surprising that some artists do not appear to seek originality and pick names that are unique. Why would anyone pick a name that is already taken by at least 80 other artists in the world (many of them still active and present online)? For current purposes, I collapse ambiguous artists (`ambiguous_artist == TRUE`) into one row per artist name.

```{r collapseartists, echo=FALSE}
lfm[, mbid := NULL]
fix <- lfm[ambiguous_artist == TRUE, ]
lfm <- lfm[ambiguous_artist == FALSE, ]

fix <- fix[,
  .(
    artist_mb = paste(unique(na.omit(artist_mb)), collapse = '; '),
    country_mb = paste(unique(na.omit(country_mb)), collapse = '; '),
    country_lastfm = paste(unique(na.omit(country_lastfm)), collapse = '; '),
    tags_mb = paste(na.omit(tags_mb), collapse = '; '),
    tags_lastfm = paste(unique(na.omit(tags_lastfm)), collapse = '; '),
    listeners_lastfm = max(listeners_lastfm, na.rm = TRUE),
    scrobbles_lastfm = max(scrobbles_lastfm, na.rm = TRUE),
    ambiguous_artist = TRUE
  ),
  by = artist_lastfm
]

lfm <- rbindlist(list(lfm, fix), fill = TRUE)
rm(fix)
```

## Unifying countries and genres

Next, I identify single country an artist comes from. The dataset contains conflicting information about artists' origin. Sometimes there is a legitimate ambiguity in what should be chosen, in other cases there is an incorrect information due to factual errors or artist duplicates. In either case, these need to be unified. Examples of problematic data:

```{r countryexample, echo=FALSE}
lfm[
  artist_lastfm %in% c('Rihanna', 'System of a Down', 'Shakira', 'Garbage', 'Gotan Project', 'Arkona'),
  .(artist_mb, country_mb, country_lastfm)
]
```

The approach is the following:

- if there is only one MusicBrainz country and it is contained in last.fm tags, this one is chosen,
- else if there is only one country in last.fm tags, this one is chosen,
- else if there is only one country in MusicBrainz tags, this one is chosen,
- to resolve conflicts in remaining entries, the most frequest country in last.fm tags is chosen.

Note, as there is no way to automatically verify whether multiple countries in last.fm tags relate to one artist or multiple artists sharing one profile, in the fourth step we might end up with an overestimate of artist popularity (only one end up being chosen, but their scrobble count comes from a shared profile).

```{r countryfix, echo=FALSE}
# unify countries
lfm[, country := NA_character_]
lfm[
  !grepl(';', country_mb) & (mapply(grepl, country_mb, country_lastfm) | is.na(country_lastfm)),
  country := country_mb
]
lfm[
  is.na(country) & !grepl(';', country_lastfm),
  country := country_lastfm
]
lfm[
  is.na(country) & !grepl(';', country_mb),
  country := country_mb
]

countries <- fread('https://raw.githubusercontent.com/ppatrzyk/kaggle/master/music/countries.csv', encoding = 'UTF-8')
patterns <- countries[,
  .(pattern = paste0(sprintf('( |;|^)%s( |;|$)', adjectival), collapse = '|')), 
  by = country
]
patterns[, pattern := paste(pattern, sprintf('( |;|^)%s( |;|$)', country), sep = '|')]
rm(countries)

first_match <- function(tags) {
  match <- patterns[,
    regexpr(pattern, tags, ignore.case = TRUE), 
    by = country
  ][V1 != -1, ][order(V1), ][1L, country]
  return(match)
}

tags <- lfm[is.na(country) & !is.na(country_lastfm), tags_lastfm]
registerDoParallel(detectCores()[1])
countries <- foreach(
  i = 1:length(tags), 
  .packages = 'data.table', 
  .combine = rbind 
) %dopar% {
  first_match(tags[i])
}

lfm[
  is.na(country) & !is.na(country_lastfm),
  country := unname(countries[, 1])
]
rm(tags, countries)
lfm[country == '', country := NA_character_]
```

In the next step, I assign genre to each artist. This step requires investigating all tags and identifying unique genres in them. This requires some aggregation over different spelling, qualifiers, etc. First, I read all tags to a separate data.table.

```{r tagexploration, echo=FALSE}
tags_lfm <- as.data.table(table(tolower(
  lfm[
    !is.na(tags_lastfm),
    unlist(strsplit(tags_lastfm, split = '; '))
  ]
)))[order(N, decreasing = TRUE), ]
```

There are `r tags_lfm[, .N]` unique tags in the dataset. `r round(100 * (tags_lfm[N == 1, .N] / tags_lfm[, .N]), 2)`% of them are used only once and only `r round(100 * (tags_lfm[N >= 10, .N] / tags_lfm[, .N]), 2)`% of them are used at least 10 times. Worcloud of the most common tags is depicted below:

```{r wordcloud, echo=FALSE, warning=FALSE}
wordcloud(
  words = tags_lfm$V1, 
  freq = tags_lfm$N, 
  min.freq = 2000, 
  scale = c(3, 0.5),
  random.order = FALSE,
  colors = viridis_pal(alpha = 1, begin = 0, end = 1, direction = 1, option = "plasma")(10)
)
```

A quick glance at tags reveals that:

- a lot of information is not related to the genre of music (e.g., *(fe)male vocalists*, *seen live*, *under 2000 listeners*, etc.),
- there are a lot of 'compound' tags -- for instance, instead of tagging a band with tags *norwegian* **and** *black metal*, people assign tags, such as *true oldschool norwegian black metal*. As a consequence, there are multiple tags referring (roughly) to the same genre. 

For instance there are `r tags_lfm[grepl('black.*metal', V1), .N]` unique *black metal* tags. Top 10:

```{r blackmetalexample, echo=FALSE}
head(tags_lfm[grepl('black.*metal', V1), .(tag = V1, count = N)], 10)
```

The approach to establishing a single genre is the following:

- for each artist, get a rank of last.fm tags,
- pick the most popular tag,
- if the picked tag is not related to genre, iteratively replace by the next tag in the rank,
- if establishing genre form last.fm tags failed, repeat the process on musicbrainz tag data.

```{r genreaggregation1, echo=FALSE}
# some of more common 'bad' tags found thorough manual inspection
filter_pattern <- paste(
  patterns[, paste(pattern, collapse = '|')],
  '^all$', "^\\d+\'*[s]?$", "live", "male", "europe", "listen", "eurovision", "^world$", "balkan", "participant", "gorgeous", "factor", "shqip", "^brasil$", "lyricist", "actress", "welsh", 'need to rate',
  "singer", "songwriter", "soundtrack", "beautiful", "awesome", "scandinav", "streamable", "duo", "radio", "century", '^african*$', 'asian*$', '^cute$', "photographer", "pianist", "francophone", 'français', 'split.*me',
  "recording", "creative", "espa", "francai", "deutsch", "portugu", "eesti", "azeri", "svenskt", "^love$", "cover", "unknown", "remix", "musician", "to.*clean.*up", "akb48", "volalist", "doujin", 'should.*be.*split',
  "funny", "humo", "super", "porn", "boobs", "science", "^muz[iy]ka$", "podcast", "test", "shit", "festival", "album", "^crap$", "^conductor$", "^guitar$", "arab", "london", "disney", "minnesota", "bogus", "guitar",
  "iskelma", "comedy", "composer", "dansk", "siberia", "norsk", "polsk", "^sex[y]*$", "epic", "parody",
  "caucasia", "albani", "suomi", "duet", "prank", "spotify", "am.*ricain", "game.*music", 'do.*not.*split',
  "^music$", "5432fun", "schlager", "oldies", "^idol$", "liedermacher", "^everything$", "^nordic$", "^amazing$", "mistagged", "^lovely$", "check", "<3", "^latin$", 'video.*game', "detroit", "mandarin", "^ost$",
  "likedis.*auto", "production.*music", "^english$", "^session$", "^england$", "kabaret", "youtube",
  'google', 'neder', '^cool$', 'novelty', 'rspiel', 'label owner', 'taiwan', 'stand.?up', 'producer', 'engineer', 'designer', '^actor$','^poet$', '^vocalist$',
  sep = '|'
)
rm(patterns)

lfm[
  !is.na(tags_lastfm),
  paste0('tag', 1:10) := tstrsplit(tolower(tags_lastfm), '; ', keep = 1:10)
]
lfm[, genre := NA_character_]
lfm[!is.na(tags_lastfm), genreset := FALSE]
for (i in 1:10) {
  lfm[genreset == FALSE, genre := get(paste0('tag', i))]
  lfm[
    genreset == FALSE, 
    genreset := !grepl(filter_pattern, genre, ignore.case = TRUE)
  ]
}
lfm[, paste0('tag', 1:10) := NULL]

lfm[, genreset := NA]
lfm[!is.na(tags_mb) & is.na(genre), genreset := FALSE]
lfm[
  !is.na(tags_mb),
  paste0('tag', 1:10) := tstrsplit(tolower(tags_mb), '; ', keep = 1:10)
]
for (i in 1:10) {
  lfm[genreset == FALSE, genre := get(paste0('tag', i))]
  lfm[
    genreset == FALSE, 
    genreset := !grepl(filter_pattern, genre, ignore.case = TRUE)
  ]
}
lfm[, paste0('tag', 1:10) := NULL]
lfm[, genre := tolower(genre)]
```

This approach yields `r lfm[, uniqueN(genre)]` unique genres, which are now aggregated. The approach here is to either find all instances of given genre via regex and unify all to one spelling (e.g., *hip-hop*), or if there are a lot of subtypes of given genre, retain some arbitrary number of most frequest ones and aggregate remainig ones into a general type (e.g., *metal*). Less frequent tags are aggregater as *other*. As I wanted to keep the number of unique genres minimal, there are some necessary arbitrary simplifications in this step (e.g., *brutal apocalyptic black gothic doom metal* becomes *black metal*).

```{r genreaggregation2, echo=FALSE}
lfm[
  grepl("hip.*hop|( |;|^)rap( |;|$)", genre) &
    !grepl("rapcore", genre),
  genre := 'hip-hop'
]

lfm[grepl("r *n *b( |$)|r.?(&|and).?b|rhyth?m.*&.*blues|rhyth?m and blues", genre), genre := 'rnb']
lfm[grepl("jazz", genre), genre := 'jazz']
lfm[grepl("trip.?hop", genre), genre := 'trip-hop']
lfm[grepl("reggae", genre), genre := 'reggae']
lfm[grepl("volksmusik|world.*music", genre), genre := 'world music']
lfm[grepl('worship|religi', genre), genre := 'worship']
lfm[grepl('soul', genre), genre := 'soul']
lfm[grepl('a.?cap', genre), genre := 'a cappella']
lfm[grepl('8.?bit', genre), genre := '8-bit']

lfm[grepl('drum.*bass', genre), genre := 'drum and bass']
lfm[grepl('dance', genre), genre := 'dance']
lfm[grepl('dis[ck]o', genre), genre := 'disco']
lfm[grepl('techno', genre), genre := 'techno']
lfm[grepl('trance', genre), genre := 'trance']
lfm[grepl('house', genre), genre := 'house']
lfm[grepl('dark.*electro', genre), genre := 'dark electro']
lfm[grepl('electro.*pop', genre), genre := 'electropop']
lfm[grepl('electro.*punk', genre), genre := 'electropunk']
lfm[
  grepl('ele[ck]tro', genre) & !(genre %in% c('dark electro', 'electropop', 'electropunk')), 
  genre := 'electronic'
]

lfm[grepl("t[h]*rash.*metal", genre), genre := 'thrash metal']
lfm[grepl("black.*metal", genre), genre := 'black metal']
lfm[grepl("death.*metal", genre), genre := 'death metal']
lfm[grepl("gothic.*metal", genre), genre := 'gothic metal']
lfm[grepl("heavy.*metal|nwobhm", genre), genre := 'heavy metal']
metal_lookup <- lfm[
  grepl("metal", genre), 
  .N, 
  by = genre
][order(N, decreasing = TRUE), ]
lfm[
  genre %in% metal_lookup[N < 40, genre],
  genre := 'metal'
]
rm(metal_lookup)

lfm[grepl('indie.*rock|rock.*indie', genre), genre := 'indie']
lfm[grepl('punk.*rock', genre), genre := 'punk']
lfm[grepl('goth', genre) & !grepl('metal', genre), genre := 'gothic rock']
lfm[grepl('rock.*rol', genre), genre := 'rock and roll']
lfm[grepl('hard.*rock', genre), genre := 'hard rock']
lfm[grepl('post.*rock', genre), genre := 'post-rock']
lfm[grepl('j.?rock', genre), genre := 'rock']
lfm[grepl('rock.*argentino', genre), genre := 'rock']

rock_lookup <- lfm[
  grepl("rock", genre), 
  .N, 
  by = genre
][order(N, decreasing = TRUE), ]
lfm[
  genre %in% rock_lookup[N < 80, genre],
  genre := 'rock'
]
rm(rock_lookup)

lfm[grepl('post.*punk', genre), genre := 'post-punk']
lfm[grepl('cold.*wave|zimna.*fala', genre), genre := 'coldwave']
lfm[grepl('dark.*wave', genre), genre := 'darkwave']
lfm[grepl('poetr|poezj', genre), genre := 'sung poetry']
lfm[grepl('folklore', genre), genre := 'folk']

punk_lookup <- lfm[
  grepl("punk", genre), 
  .N, 
  by = genre
 ][order(N, decreasing = TRUE), ]
lfm[
  genre %in% punk_lookup[N < 20, genre],
  genre := 'punk'
]
rm(punk_lookup)

lfm[
  grepl('pop', genre) & !(genre %in% c('electropop', 'synthpop')), 
  genre := 'pop'
]

all_lookup <- lfm[,
  .N, 
  by = genre
 ][order(N, decreasing = TRUE), ]
lfm[
  genre %in% all_lookup[N < 30, genre],
  genre := NA_character_
]
rm(all_lookup)
lfm[genre == '', genre := NA_character_]
```

```{r nameunification, echo=FALSE}
setnames(lfm, 'artist_lastfm', 'artist')
lfm[is.na(artist), artist := artist_mb]
lfm[, c("artist_mb", "genreset", "country_mb", "country_lastfm", "tags_mb", "tags_lastfm", "ambiguous_artist") := NULL]
```

# Popularity distribution

Having set up the data, let's see who are the top 100 artists (sorted by the number of listeners):

```{r toptenartist, echo=FALSE}
datatable(lfm[1:100, ])
```

The data is extremely skewed -- the vast majority of artists have a very small number of listeners and scrobbles while there are few ones with extraordinary popularity.

## ECDF: listeners per artist

Note that, x axis is pruned at 50,000 listeners -- this is approximately `r paste(round(100 * 50000 / lfm[1L, listeners_lastfm], 2), "%")` of how many listeners the top artist in dataset (i.e., Coldplay) has. 

```{r popularitycdf, echo=FALSE}
lfm[is.na(listeners_lastfm), listeners_lastfm := 0]
lfm[is.na(scrobbles_lastfm), scrobbles_lastfm := 0]
get_percentile <- lfm[, ecdf(listeners_lastfm)]
lfm[, percentile := get_percentile(listeners_lastfm)]

ecdf_plot <- ggplot(
  lfm,
  aes(x = listeners_lastfm, y = percentile)
) +
  coord_cartesian(
    xlim = c(0, (10^5)/2)
    # ylim = c(0, 1)
  ) +
  scale_x_continuous(
    breaks = seq(0, (10^5)/2, by = 1.25 * 10^4),
    labels = display_values
  ) +
  geom_line(
    color = "black",
    size = 1
  ) +
  labs(
    title = "Empirical CDF",
    x = "Number of listeners",
    y = "Fraction of data"
  ) +
  annotate(
    "segment",
    x = 0,
    xend = 1.3 * 10^6,
    y = 1,
    yend = 1,
    size = 0.5,
    linetype = "dashed"
  ) +
  theme_bw()

print(ecdf_plot)
```

## Histogram: listeners per artist

Note that artist count is displayed on a 4th-root-transformed scale.

```{r popularityhistogram, echo=FALSE}
listeners_hist <- ggplot(
  lfm, aes(x = listeners_lastfm)
) + 
  geom_histogram(
    binwidth = 10^5,
    fill = '#ED7953FF'
  ) +
  scale_y_continuous(
    trans = scales::trans_new(
      name = "test_tans", 
      transform = function(x) {x^(1/4)}, 
      inverse = function(x) {x^4}
    ),
    breaks = c(1, 10, 10^4, 10^5, 10^6),
    labels = display_values
  ) +
  scale_x_continuous(
    labels = display_values
  ) +
  labs(
    title = "Histogram of number of listeners per artist",
    x = "Number of listeners",
    y = "Artists (4th root transform)"
  ) +
  theme_bw()

print(listeners_hist)
```

## Listeners vs scrobbles scatter

Another way to look at this is to see a scatterplot of listeners vs scrobbles for all artists in the world.

```{r hexbinscatter, echo=FALSE}
artist_label <- lfm[
  order(scrobbles_lastfm, decreasing = TRUE), 
][
  1:8,
  .(artist, x = listeners_lastfm, y = scrobbles_lastfm)
]

hexbin_scatter <- ggplot(
  lfm, aes(x = listeners_lastfm, y = scrobbles_lastfm)
) +
  stat_binhex(
    show.legend = TRUE,
    bins = 60,
    color = "black"
  ) +
  scale_fill_gradientn(
    name = "Number of artists\n(log scale)", trans = "log", 
    colors = viridis_pal(alpha = 1, begin = 0, end = 1, direction = -1, option = "plasma")(10),
    breaks = c(1, 100, 8000, 590000), labels = display_values
  ) +
  scale_x_continuous(
    breaks = seq(0, 5 * 10^6, by = 10^6),
    labels = display_values
  ) + 
  scale_y_continuous(
    breaks = seq(0, 500 * 10^6, by = 10^8),
    labels = display_values
  ) +
  labs(
    title = "Artist listeners and scrobbles",
    x = "Number of listeners",
    y = "Number of scrobbles"
  ) + 
  annotate(
    "text",
    x = artist_label$x,
    y = artist_label$y,
    label = artist_label$artist,
    size = 3.5
  ) +
  annotate(
    "curve",
    x = 0,
    xend = 6 * -10^4,
    y = 10^8, 
    yend = 5 * 10^6,
    size = 1,
    arrow = arrow(length = unit(0.2, "cm"))
  ) +
  annotate(
    "label", 
    x = 0, 
    y = 10^8, 
    label = "98.7% of all\nartists",
    # hardcoded here, for verification:
    # ggplot_build(hexbin_scatter)$data[[1]]
    size = 3.5
  ) +
  theme_bw()

print(hexbin_scatter)
```

## Scrobbles concentration

```{r scrobblescumsum, echo = FALSE}
scrobbles <- lfm[,
  .(artist, scrobbles_lastfm)
][
  order(scrobbles_lastfm, decreasing = TRUE), 
]
scrobbles[, total := sum(as.numeric(scrobbles_lastfm))]
scrobbles[, cumulative := cumsum(as.numeric(scrobbles_lastfm))]
scrobbles[, proportion := cumulative / total]
scrobbles[, distance_half := abs(proportion - 0.5)]
scrobbles[, distance_90 := abs(proportion - 0.9)]
```

All world scrobbles (i.e., track plays) are concentrated on very limited set of artists. Specifically:

- 50% of the world's scrobbles is generated by `r sprintf('%s (%s%%)', scrobbles[, which.min(distance_half)], round(100 * scrobbles[, which.min(distance_half)] / scrobbles[, .N], 2))` artists,
- 90% of all scrobbles are generated by `r sprintf('%s (%s%%)', scrobbles[, which.min(distance_90)], round(100 * scrobbles[, which.min(distance_90)] / scrobbles[, .N], 2))` artists.

Visualizing this:

```{r propviz, echo=FALSE, fig.height=2}
rectangles <- data.table(
  x = 0,
  y = rep(c(1, 2), 2),
  width = c(10, 10, 9, (10 * scrobbles[, which.min(distance_90)] / scrobbles[, .N])),
  type = rep(c('full', 'prop'), each = 2)
)

prop_rectangles <- ggplot(rectangles, aes(xmin = x, xmax = width, ymin = y, ymax = y + 0.5)) +
  geom_rect(aes(fill = type), color = "grey50", show.legend = FALSE) +
  scale_fill_manual(values = c('grey', '#ED7953FF')) +
  annotate(
    "text", x = 5, y = c(1.6, 2.6), 
    label = c("Generate this proportion of scrobbles:", "This proportion of artists:")
  ) +
  theme_void()

prop_rectangles
```

# Summarizing data

Summarizing data by country and genre poses the problem of handling missing values. In the dataset, any of these two variables can be missing -- hence, prior to proceeding we need to check if the presence of missing values is random. Because data is sensitive to outliers (very succesful bands), here I observe aggregated number of scrobbles in missing entries rather than their raw count. First, looking at a country-level, here is a summary of proportion of music with missing genre. In other words, if we a take a single country, for what proportion of music we cannot tell what genre is it.

```{r missinggenrecheck, echo=FALSE}
country_test <- dcast.data.table(
  lfm[,
    sum(as.numeric(scrobbles_lastfm)), 
    by = .(
      country,
      missing_genre = ifelse(is.na(genre), 'missing', 'present')
    )
  ],
  country ~ missing_genre,
  value.var = 'V1'
)
country_test[is.na(country_test)] <- 0
country_test[, prop_missing := round(missing / (missing + present), 2)]
country_test[, summary(prop_missing)]
```

Second, a summary of proprotion of missing countries on a genre level. In other words, if we a take a single genre, for what proportion of music we cannot tell the country of origin.

```{r missingcountrycheck, echo=FALSE}
genre_test <- dcast.data.table(
  lfm[,
    sum(as.numeric(scrobbles_lastfm)), 
    by = .(
      genre,
      missing_country = ifelse(is.na(country), 'missing', 'present')
    )
  ],
  genre ~ missing_country,
  value.var = 'V1'
)
genre_test[is.na(genre_test)] <- 0
genre_test[, prop_missing := round(missing / (missing + present), 2)]
genre_test[, summary(prop_missing)]
```

Inspection reveals that it is not possible to reliably analyze genre vs. country associations for less popular genres or smaller countries, as non-negligible proportion of data is missing. However, for bigger categories, there is only about ~5% of data in other category missing and this number is stable across diffrent categories (countires or genres). In subsequent analysis, I take subset of data where both country and genre information is present.

```{r summarysubsets, echo=FALSE}
genres <- lfm[
  !(is.na(genre) | is.na(country)),
  .(
    artists = .N,
    topartist = artist[1],
    topartist_country = country[1],
    allscrobbles = sum(as.numeric(scrobbles_lastfm))
  ),
  by = genre
][
  order(allscrobbles, decreasing = TRUE)
]

country_all <- lfm[
  !(is.na(genre) | is.na(country)),
  .(
    artists = .N,
    topartist = artist[1],
    topartist_genre = genre[1],
    allscrobbles = sum(as.numeric(scrobbles_lastfm))
  ),
  by = country
][
  order(allscrobbles, decreasing = TRUE)
]
country_all[, percent_of_all_scrobbles := round(100 * allscrobbles /  sum(as.numeric(allscrobbles)), 2)]

genre_country <- lfm[
  !(is.na(genre) | is.na(country)),
  .(
    # artists = .N,
    allscrobbles = sum(as.numeric(scrobbles_lastfm))
  ),
  by = .(genre, country)
]

genre_country <- merge(
  genre_country, 
  genres[, .(genre, genretotal = allscrobbles)],
  all.x = TRUE,
  by = 'genre'
)
genre_country <- merge(
  genre_country, 
  country_all[, .(country, countrytotal = allscrobbles)],
  all.x = TRUE,
  by = 'country'
)
genre_country[, genre_perc := round(100 * allscrobbles / genretotal, 2)]
genre_country[, country_perc := round(100 * allscrobbles / countrytotal, 2)]
```

# Top Genres

What are the top musical genres in the dataset?

## All scrobbles by genre

A scatterplot of number of artists and aggregated number of scrobbles for each genre. Top genres are annotated.

```{r topgenres, echo=FALSE, warning=FALSE}
genres_top <- genres[
  allscrobbles >= 500 * 10^6,
  .(genre, artists, allscrobbles)
]

genres_scatter <- ggplot(genres_top, aes(artists, allscrobbles)) +
  geom_point(shape = 21, fill = '#ED7953FF', color = 'black', stroke = 1, size = 4, alpha = 0.4) +
  scale_x_continuous(labels = display_values) + 
  scale_y_log10(labels = display_values) +
  labs(
    title = "Artists and total scrobbles by genre",
    x = "Number of artists",
    y = "Aggregated scrobbles (Log scale)"
  ) + 
  annotate(
    'text',
    x = genres_top[allscrobbles > (1.3 * 10^9) | artists > 5000, artists],
    y = genres_top[allscrobbles > (1.3 * 10^9) | artists > 5000, allscrobbles],
    label = genres_top[allscrobbles > (1.3 * 10^9) | artists > 5000, genre],
    size = 3.5
  ) +
  theme_bw()

genres_scatter
```

# Geography of music production

Where the top music is produced.

```{r leafletbase, echo=FALSE}
# base map to be re-used
opts <- leafletOptions(minZoom = 2, maxZoom = 5, worldCopyJump = TRUE)
world_map <- leaflet(options = opts) %>%
  setView(lat = 50, lng = 15, zoom = 4) %>%
  addMapPane("background_map", zIndex = 410) %>%
  addMapPane("polygons", zIndex = 420) %>%
  addMapPane("labels", zIndex = 430) %>%
  addProviderTiles(
    "Esri.WorldTerrain", #"Stamen.TonerBackground"
    providerTileOptions(zIndex=410)
  )

countries <- fread('https://raw.githubusercontent.com/ppatrzyk/kaggle/master/music/country_codes.csv')
world <- geojson_read("https://raw.githubusercontent.com/ppatrzyk/kaggle/master/music/ne_110m_admin_0_countries.geo.json", what = "sp")
world@data$index <- 1:nrow(world@data)
world@data <- merge(
  world@data[, c('index', "iso_a3")], 
  countries, 
  by.x = "iso_a3",
  by.y = "A3",
  all.x = TRUE
)
world@data <- world@data[order(world@data$index), ]
centers <- data.frame(rgeos::gCentroid(world, byid = TRUE))
# manually fix strange centers in some countries
centers[which(world@data$country == "France"), "x"] <- 2.32259
centers[which(world@data$country == "France"), "y"] <- 46.67595
centers[which(world@data$country == "Norway"), "x"] <- 7.56845
centers[which(world@data$country == "Norway"), "y"] <- 60.73180
centers[which(world@data$country == "United States"), "x"] <- -101.44575
centers[which(world@data$country == "United States"), "y"] <- 39.75996
centers[which(world@data$country == "United Kingdom"), "x"] <- -0.93118
centers[which(world@data$country == "United Kingdom"), "y"] <- 52.00022
centers[which(world@data$country == "Bosnia and Herzegovina"), "y"] <- 43.4
```

## Country ranking

First, let's examine the ranking of all counties by the total number of scrobbles their artists produce. 

```{r countryranking, echo=FALSE}
datatable(
  country_all[
    !is.na(country) & country != '' & allscrobbles > 10^5,
  ][
    order(allscrobbles, decreasing = TRUE),
  ]
)
```

## Country barplot

```{r countrybarplot, echo=FALSE}
country_bar_data <- rbindlist(list(
  country_all[1:10, .(country, percent_of_all_scrobbles)],
  country_all[11:.N, .('other', percent_of_all_scrobbles = sum(percent_of_all_scrobbles))]
))
country_bar_data[, country := factor(country, levels = rev(country))]

country_bar <- ggplot(
  country_bar_data,
  aes(country, percent_of_all_scrobbles)
) +
  geom_col(
    fill = '#ED7953FF'
  ) +
  coord_flip() +
  labs(
    title = "Where does the most listened music come from?",
    x = "Music origin",
    y = "Percent of world's listened to music"
  ) +
  theme_bw()

country_bar
```

## Music production map

Countries on map are colored by the total number of scrobbles produced (empirical quantile) labeled by the top artists coming from there. Click on a country to get its name and actual scrobble count.

```{r scrobblesmap, echo=FALSE}
topartist_data <- world
topartist_data@data <- merge(
  topartist_data@data,
  country_all[,
    .(
      country, 
      topartist, 
      allscrobbles, 
      popup = paste0(
        country, ': ', display_values(allscrobbles), ' scrobbles'
      )
    )
  ],
  all.x = TRUE,
  by = 'country'
)
topartist_data@data <- topartist_data@data[order(topartist_data@data$index), ]

quant_pal <- colorQuantile(
  viridis_pal(alpha = 1, begin = 0, end = 1, direction = 1, option = "plasma")(10), 
  topartist_data@data$allscrobbles, 
  n = 10
)

totalscrobbles_map <- addPolygons(
  world_map, data = topartist_data,
  color = "black", opacity = 1, weight = 1,
  smoothFactor = 0.2, fillOpacity = 0.6, fillColor = ~quant_pal(allscrobbles),
  popup = topartist_data@data$popup,
  options = pathOptions(pane = "polygons")
) %>%
  leaflet::addLegend(
    position = "bottomright", values = topartist_data@data$allscrobbles,
    title = "Total scrobbles", pal = quant_pal, opacity = 1
  ) %>%
  addLabelOnlyMarkers(
    data = centers, lng = ~x, lat = ~y, 
    label = topartist_data@data$topartist,
    labelOptions = labelOptions(
      noHide = TRUE, 
      direction = "center", 
      style = list(
        "background" = "rgba(240, 240, 240, 0.4)",
        "font-weight" = "bold",
        "border" = "none",
        "box-shadow" = "none"
      )
    ),
    options = pathOptions(pane = "labels")
  )

totalscrobbles_map
```

## Where USA and UK do not dominate

When looking at raw data, almost half (`r country_all[country == 'United States', percent_of_all_scrobbles]`%) of the songs listened to are made by artists coming from the United States. The next country, United Kingdom, generates `r country_all[country == 'United Kingdom', percent_of_all_scrobbles]`% of all scrobbles. This skew is even more apparent when looking at the origin of the top 1000 artists in the world -- Only `r lfm[1:1000, ][!(country %in% c('United States', 'United Kingdom')), .N]` of the top 1k artists do not come from either US or UK.

Having confirmed and quantified what is widely know about the mainstream music, let's investigate whether this dominance is present for all genres.

There is some variablity in country rankings when we look at specific genres rather that the entire data:

```{r bygenreranks, echo=FALSE, fig.height=240}
top_genres <- genres[1:96, genre]
genres_rank_data <- replicate(96, data.table())
for (i in 1:length(top_genres)) {
  current_genre <- top_genres[i]
  current_subset <- genre_country[
    order(genre_perc, decreasing = TRUE),
  ][
    genre == current_genre, 
  ]
  topgenres <- rbindlist(list(
    current_subset[1:10, .(genre = current_genre, country, genre_perc)],
    current_subset[11:.N, .(genre = current_genre, 'other', genre_perc = sum(genre_perc))]
  ))
  genres_rank_data[[i]] <- topgenres
}
genres_rank_data <- rbindlist(genres_rank_data)
genres_rank_data[, factor_order := factor(1:.N, levels = .N:1)]
genres_rank_data[, genre := factor(genre, levels = unique(genre))]

genres_rank_bar <- ggplot(
  genres_rank_data,
  aes(factor_order, genre_perc)
) +
  geom_col(
    fill = '#ED7953FF'
  ) +
  scale_x_discrete(
    labels = function(factor_no) genres_rank_data[factor_order %in% factor_no, rev(country)]
  ) +
  coord_flip() +
  labs(
    title = "Country ranking for specific genres",
    x = "Country",
    y = "Percent of produced music"
  ) +
  theme_bw() +
  theme(
    axis.text.y = element_text(color = "black", size = 12)
  ) +
  facet_wrap(
    vars(genre), 
    scales = "free",
    nrow = 48,
    ncol = 2
  )

genres_rank_bar
```

List of genres where the top producer (aggregated scrobbles) is neither USA nor UK or the top artist is not from there: 

```{r topnotus, echo=FALSE}
byscrobbles <- genre_country[
  order(genre_perc, decreasing = TRUE),
  .(
    top_producer = country[1],
    perc_all = genre_perc[1],
    genretotal = genretotal[1]
  ),
  by = genre
][,
  .(genre, top_producer, perc_all, genretotal)
]

datatable(
  merge(
    genres[, .(genre, topartist, topartist_country)],
    byscrobbles,
    by = 'genre'
  )[
    !(topartist_country %in% c('United States', 'United Kingdom')) |
      !(top_producer %in% c('United States', 'United Kingdom')),
  ][
    genretotal > 10^7
  ][
    order(genretotal, decreasing = TRUE),
    .(genre, top_producer, topartist_country, topartist)
  ]
)
```

# Genre in country-level production

Another interesting question is to observe relative popularity of specific genres within the entire music produced by specific countries. In other words, here I isolate top 24 countries and observe which genres make up their music export.

```{r, relativeproduction, echo=FALSE, fig.height=60}
top_countries <- country_all[1:24, country]
genres_within_data <- replicate(24, data.table())
for (i in 1:length(top_countries)) {
  current_country <- top_countries[i]
  current_subset <- genre_country[
    order(country_perc, decreasing = TRUE),
  ][
    country == current_country, 
  ]
  topgenres <- rbindlist(list(
    current_subset[1:10, .(country = current_country, genre, country_perc)],
    current_subset[11:.N, .(country = current_country, 'other', country_perc = sum(country_perc))]
  ))
  genres_within_data[[i]] <- topgenres
}
genres_within_data <- rbindlist(genres_within_data)
genres_within_data[, factor_order := factor(1:.N, levels = .N:1)]
genres_within_data[, country := factor(country, levels = unique(country))]

genres_within_bar <- ggplot(
  genres_within_data,
  aes(factor_order, country_perc)
) +
  geom_col(
    fill = '#ED7953FF'
  ) +
  scale_x_discrete(
    labels = function(factor_no) genres_within_data[factor_order %in% factor_no, rev(genre)]
  ) +
  coord_flip() +
  labs(
    title = "Within country popularity",
    x = "Genre",
    y = "Percent of produced music"
  ) +
  theme_bw() +
  theme(
    axis.text.y = element_text(color = "black", size = 12)
  ) +
  facet_wrap(
    vars(country), 
    scales = "free",
    nrow = 12,
    ncol = 2
  )

genres_within_bar
```

# Conclusion

Last.fm users do not constitute representative sample of world's music listeners. Hence, the number of scrobbles is an imperfect proxy to the number of times given artist has been listened to. 